---
title: "Tidy Tuesday - Shakespeare"
description: |
  Analysis of Shakespeare data from TidyTuesday
author: Rowan Norenberg
date: October 2, 2025
format:
  html:
    code-fold: true
    code-summary: "Show the code"
---

I've always loved Shakespeare, with Macbeth, Hamlet, and Julius Cesar in particular holding a special place in my heart.

```{r message=FALSE, warning=FALSE}
#| echo: true

#| echo: false
library(tidyverse)
library(ggplot2)
library(dplyr)
library(stringr)
library(forcats)


hamlet <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-09-17/hamlet.csv')
macbeth <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-09-17/macbeth.csv')
romeo_juliet <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-09-17/romeo_juliet.csv')

```

```{r}
#| echo: true
#| message: false
#| warning: false


tragedies <- bind_rows(
  hamlet |> mutate(play = "Hamlet"),
  macbeth |> mutate(play = "Macbeth"),
  romeo_juliet |> mutate(play = "Romeo and Juliet")
)


tragedies_ghosts <- tragedies |>
  filter(str_detect(dialogue, "\\b(ghost|witch|spirit)\\b")) |>
  select(play, dialogue) |>
  mutate(supernatural = str_extract(str_to_lower(dialogue), "\\b(ghost|witch|spirit)\\b" )) |>
  count(play, supernatural, name = "num_mentions")



ggplot(tragedies_ghosts, aes(x = play, y = num_mentions, fill = supernatural)) +
  geom_col(position = "dodge") +
  labs(
    title = "Mentions of 'Ghost', 'Witch', and 'Spirit' in Shakespeare's Tragedies",
    x = "Play",
    y = "Number of Mentions",
    fill = "" 
    ) + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 16, face = "bold"))


```

This graph shows that Hamlet has a great deal many more mentions of ghosts, witches, and spirits than the other two plays. This indicates that ghosts, witches, and spirits are more relevant to the plot of Hamlet than to the other two, which makes sense because Hamlet is frequently seeing apparitions of his dead father and this is central to the plot.

```{r}
#| echo: true
#| message: false
#| warning: false


tragedies_exclaim <- tragedies |>
  filter(str_detect(dialogue, "(?i)\\b(Hamlet|Macbeth|Romeo|Juliet)!+")) |>
  mutate(exclamations = str_extract(dialogue, "(?i)\\b(Hamlet|Macbeth|Romeo|Juliet)!+")) |>
  count(play, exclamations, name = "num_exclaim") |>
  mutate(exclamations = fct_relevel(exclamations, c("Hamlet!", "Macbeth!", "Romeo!", "Juliet!")))



ggplot(tragedies_exclaim, aes(x = exclamations, y = num_exclaim, fill = exclamations)) +
  geom_col(position = "dodge") +
  labs(
    title = "Number of times the main character's name is exclaimed",
    x = "Play",
    y = "Number of Mentions",
    fill = "" 
    ) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 16, face = "bold"))


```

This graph shows that out of the main characters from each play, Romeo is the one who gets yelled at (or about) the most.

```{r}
#| echo: true
#| message: false
#| warning: false

tragedies_exclaimers <- tragedies |>
  filter(str_detect(dialogue, "(?i)\\b(Hamlet|Macbeth|Romeo|Juliet)!+")) |> 
  mutate(
    exclamations = str_extract(dialogue, "(?i)\\b(Hamlet|Macbeth|Romeo|Juliet)!+"),
    exclamations = str_to_title(exclamations)  
  ) |>
  count(play, character, exclamations, name = "num_exclaim") |>
  arrange(desc(num_exclaim))


top_exclaimers <- tragedies_exclaimers |> 
  slice_max(num_exclaim, n = 10)

ggplot(top_exclaimers, aes(
  x = reorder(character, num_exclaim),
  y = num_exclaim,
  fill = exclamations
)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Characters Who Shout Names the Most",
    subtitle = "Colored by which name they exclaim",
    x = "Character",
    y = "Number of Exclamations",
    fill = "Name Shouted"
  ) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 16, face = "bold"))


```

And this graph shows that not all of those times Romeo is yelled at are from Juliet's famous line! Turns out Romeo is yelled at (or about) by a good variety of folks.

```{r}
#| echo: true
#| message: false
#| warning: false


tragedies_exclaimers <- tragedies |>
  filter(str_detect(dialogue, "(?i)!+")) |> 
  mutate(
    exclamations = str_extract(dialogue, "(?i)!+"),
    exclamations = str_to_title(exclamations)  
  ) |>
  count(play, character, exclamations, name = "num_exclaim") |>
  arrange(desc(num_exclaim))

# Optionally, focus on the top 10 loudest shouters
top_exclaimers <- tragedies_exclaimers |> 
  slice_max(num_exclaim, n = 10)

ggplot(top_exclaimers, aes(
  x = reorder(character, num_exclaim),
  y = num_exclaim,
  fill = play
)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Criers",
    subtitle = "Characters in Shakespeare's Tragedies who Exclaim the Most",
    x = "Character",
    y = "Number of Exclamations",
    fill = ""
  ) +
  theme_minimal() + 
  theme(plot.title = element_text(size = 16, face = "bold"))

```

Finally, this graph tells us that despite not being yelled at nearly as much as Romeo is, Hamlet yells a lot. Perhaps this is because of all of those ghosts, spirits, and witches.

Below is a graph for each play showing who has the most lines. These line counts graphs tell you which character is the "main character", which, unsurprisingly, are also the title characters. But just in case.

```{r}
#| echo: true
#| message: false
#| warning: false

hamlet_counts <- hamlet |>
  filter(str_extract(character, "stage", ignore.case = TRUE)) |> 
  count(character, name = "num_lines") |>
  mutate(play = "Hamlet") |>
  arrange(desc(num_lines)) |>
  slice_head(n = 20)

  
  
ggplot(hamlet_counts, aes(x = reorder(character, num_lines), y = num_lines)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Hamlet: Number of Lines",
    x = "Named Character",
    y = "Total Number of Lines"
  ) +
  theme_minimal(base_size = 14) +  # bigger text
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    axis.text.y = element_text(size = 10)
  ) + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 16, face = "bold"))



```

```{r}
#| echo: true
#| message: false
#| warning: false
macbeth_counts <- macbeth |>
  filter(!grepl("stage", character, ignore.case = TRUE)) |> 
  count(character, name = "num_lines") |>
  mutate(play = "Macbeth") |>
  arrange(desc(num_lines)) |>
  slice_head(n = 20)

  
  
ggplot(macbeth_counts, aes(x = reorder(character, num_lines), y = num_lines)) +
  geom_col(fill = "goldenrod") +
  coord_flip() +
  labs(
    title = "Macbeth: Number of Lines",
    x = "Named Character",
    y = "Total Number of Lines"
  ) +
  theme_minimal(base_size = 14) +  # bigger text
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    axis.text.y = element_text(size = 10)
  ) + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 16, face = "bold"))


```

```{r}
#| echo: true
#| message: false
#| warning: false
romeo_juliet_counts <- romeo_juliet |>
  filter(!grepl("stage", character, ignore.case = TRUE)) |> 
  count(character, name = "num_lines") |>
  mutate(play = "Romeo and Juliet") |>
  arrange(desc(num_lines)) |>
  slice_head(n = 20)

  
  
ggplot(romeo_juliet_counts, aes(x = reorder(character, num_lines), y = num_lines)) +
  geom_col(fill = "maroon") +
  coord_flip() +
  labs(
    title = "Romeo and Juliet: Number of Lines",
    x = "Named Character",
    y = "Total Number of Lines"
  ) +
  theme_minimal(base_size = 14) +  # bigger text
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    axis.text.y = element_text(size = 10)
  ) + 
  theme_minimal() + 
  theme(plot.title = element_text(size = 16, face = "bold"))


```

*The Complete Works of William Shakespeare*. (n.d.). Retrieved December 10, 2025, from <https://shakespeare.mit.edu/>

*TheMITTech/shakespeare: The Complete Works of William Shakespeare hosted at http://shakespeare.mit.edu/*. (n.d.). Retrieved December 10, 2025, from <https://github.com/TheMITTech/shakespeare>

*Tidytuesday/data/2024/2024-12-17 at main rfordatascience/tidytuesday*. (n.d.). GitHub. Retrieved December 10, 2025, from <https://github.com/rfordatascience/tidytuesday/tree/main/data/2024/2024-12-17>
